---
title: "Projet"
author: "MOUDILA Marcel"
date: "`r Sys.Date()`"
output: beamer_presentation
---

**Abstract**: This collection of data is part of the RNA-Seq (HiSeq) PANCAN data set, it is a random extraction of gene expressions of patients having different types of tumor: BRCA, KIRC, COAD, LUAD and PRAD.

|                                |                            |
|--------------------------------|----------------------------|
| **Data Set Characteristics:**  | Multivariate               |
| **Attribute Characteristics:** | Real                       |
| **Associated Tasks:**          | Classification, Clustering |
| **Number of Instances:**       | 801                        |
| **Number of Attributes:**      | 20531                      |
| **Missing Values?**            | N/A                        |

**Data Set Information:**

Samples (instances) are stored row-wise. Variables (attributes) of each sample are RNA-Seq gene expression levels measured by illumina HiSeq platform.

**Attribute Information:**

A dummy name (gene_XX) is given to each attribute. The attributes are ordered consitently with the original submission.

```{r include=FALSE}
library(data.table)
data = fread("/home/moudilamarcel/Bureau/data.csv")
labels = fread("/home/moudilamarcel/Bureau/labels.csv", header = TRUE)
data = as.data.frame(data[,-1])
labels = as.data.frame(labels[,-1])

# quelques manipulations
names(labels) = "Class"
labels$Class = as.factor(labels$Class)
df = cbind.data.frame(labels,data)
```

Steps :

1.  Data analysis

2.  Anova's test and Bonferroni's test to  reduce the size 

3.  missing values analysis

4.  training 1: logistic regression

5.  training 2: random forest

6.  training 3: linear discriminant analysis

# Data analysis

Our dataset has  big size (801,20532), no missing values. The target is of type categorical, this is classification's problem.

|          | BRCA | COAD | KIRC | LUAD | PRAD |
|----------|------|------|------|------|------|
| Count    | 300  | 78   | 146  | 141  | 136  |
|          |      |      |      |      |      |

: Table of count

```{r}
XtrainFinal <- read.csv2("~/MCAD Marcel/Stat de grande dimension/projet/XtrainFinal.csv")
X_train = XtrainFinal[,2:ncol(XtrainFinal)]

XtestFinal <- read.csv2("~/MCAD Marcel/Stat de grande dimension/projet/XtestFinal.csv")
X_test = XtestFinal[,2:ncol(XtestFinal)]

```


```{r}
ytrainFinal <- read.csv("~/MCAD Marcel/Stat de grande dimension/projet/ytrainFinal.csv", sep=";")
y_train = ytrainFinal[,2]

ytestFinal <- read.csv("~/MCAD Marcel/Stat de grande dimension/projet/ytestFinal.csv", sep=";")
y_test = ytestFinal[,2]
```


# training 

## Logistic regression

```{r echo=TRUE}
# cross-validation avec 10 folds
fitControl <- trainControl(method = "cv", number = 10)
```

```{r echo=TRUE}
# régression logistique multinomiale

regLogistique <- train(X_train, y_train, 
                 method = "multinom",
                 trControl = fitControl,
                 metric = "Accuracy",
                 trace = FALSE)
```                 


```{r}
regLogistique
```

```{r}
pred_regLogistique <- predict(regLogistique, X_test)
```


```{r}
# score on testing dataset
postResample(pred = pred_regLogistique, obs = y_test)
```

## Random Forest 

```{r echo=TRUE}
# random forest
randomForest <- train(X_train, y_train, 
                 method = "rf",
                 trControl = fitControl,
                 metric = "Accuracy")
```


```{r}
randomForest
```

```{r}
pred_randomForest <- predict(randomForest, X_test)
```


```{r}
# score on testing dataset
postResample(pred = pred_randomForest, obs = y_test)
```

## linear discriminant analysis

```{r include=FALSE}
ADL <- train(X_train, y_train, 
                 method = "lda2",
                 trControl = fitControl,
                 metric = "Accuracy")
```


```{r}
ADL
```

```{r}
pred_ADL <- predict(ADL, X_test)
```


```{r}
# score on testing dataset 
postResample(pred = pred_ADL, obs = y_test)
```

# The best: Logistic regression

# Other method : SPLSDA (my preference) 

```{r include=FALSE}
library(MASS)
library(rgl)
library(lattice)
library(pheatmap)
library(mixOmics)
```

```{r include=FALSE}
X_train = train[,2:18379]
X_test = test[,2:18379]
X_train <- X_train[,-highlyCorDescr]
X_test <- X_test[,-highlyCorDescr]

mat = data.matrix(X_train,rownames.force = NA)

tumeursplsda <-splsda(mat, y_train, ncomp = 3, keepX = rep(10,3))
```

```{r include=FALSE}
ind <- which(tumeursplsda$loadings$X[,]!=0,arr.ind=TRUE)
res <- tumeursplsda$loadings$X[ind[,1],]
rownames(res) <- ind[,1]
```

```{r}
plotIndiv(tumeursplsda, ind.names = y_train, legend = TRUE, ellipse =TRUE)
```

```{r include=FALSE}
tumeur.5foldda <- perf(tumeursplsda, validation = "Mfold", folds = 5,
dist = "all", auc = TRUE)
```

```{r echo=TRUE}
tumeur.5foldda$auc
```

