---
title: "TP1 stat de grande dimension"
author: "Marcel MOUDILA, GANTOUNKPO mafoya Jean Noel"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
# 2 Multiple testing and p-values

## 1) Simulate a vector of binary variable Y of size n = 50, 100, 200
```{r}
n<-50
prob <- 0.4
Y<-(runif(n,0,1)<prob)*1
#print(Y)
length(Y)
```
## 2) Simulate a matrix of quantitatives covariates X of size n = 50 and p = 100 that is not linked to Y
```{r}
p <- 100
X <- matrix(rnorm(n*p,0,1), ncol=p)
#print(X)
dim(X)
```
## 3) What is the appropriate test to verify if one covariate is linked to Y ?

the appropriate test to verify if one covariate is linked to Y is Student's t-Test

## 4) Perform the test for each covariate and save the p-values in a vector. What is the min and the max of the p-values ? Plot the histogram. Comment.
```{r}
pvalue <- c()
for (i in 1:p){
  df = data.frame(column1 = Y, column2 = X[,i])
  Test = t.test(df$column2[which(df$column1==1)],
                df$column2[which(df$column1==0)],
                level = 0.95)
  pvalue <- c(pvalue,Test$p.value)
}
print(summary(pvalue))
hist(pvalue,20,main="p-value test de lien entre X et Y")
```

The covoriates are not linked with Y normally, but there is the tests with p-value < 0.05, so there is the False Positive Significant tests.

## 5) Compute the number of False Positive Significant tests for a level alpha
```{r}
FP = length(which(pvalue < 0.05)) # false positive
print(FP)
```

## 6) Compute the empirical type I error and compare to alpha 

The empirical type I error is as the number of False Positive Significant tests

```{r}
empirical_type1_error = FP/p
print(empirical_type1_error)
```


## 7) Do it again for p = 1000 and p = 10000

## 8) We want now to simulate a matrix X of covariates of size n = 50 and p1 = 100 that are linked with Y by the following way

## a) matrix X1, and min, max, histogram of p-value's vector and number of Significant tests and the empirical power of the test
```{r}
p1 = 100
vect = c()
for (j in 1:p1){
Z <- (runif(n,0,1)<0.7)*1
obs <- Z*rnorm(n,3*Y,1) + (1-Z)*rnorm(n,0,1) #obs : n realisations de X_j
vect = c(vect,obs)
}
X1 <- matrix(vect,ncol=p1)
#print(X1)
print(dim(X1))


pvalue <- c()
for (i in 1:p1){
  df = data.frame(column1 = Y, column2 = X1[,i])
  Test = t.test(df$column2[which(df$column1==1)],
                df$column2[which(df$column1==0)],
                level = 0.95)
  pvalue <- c(pvalue,Test$p.value)
}
pv1 = pvalue
print(summary(pv1))
hist(pv1,20,main="p-value test colonnes de X1 et Y",col = "blue")
print(length(which(pv1 < 0.05))) # number of significant tests


```
The covariates are linked with Y normally. Exactly, we see it because most p-value < 0.05

## b) other matrice X2, and min, max, histogram of p-value's vector and number of Significant tests and the empirical power of the test

```{r}
vect = c()
for (j in 1:p1){
Z <- (runif(n,0,1)<0.7)*1
obs <- Z*rnorm(n,2*Y,1) + (1-Z)*rnorm(n,0,1) #obs : n realisations de X_j
vect = c(vect,obs)
}
X2 <- matrix(vect,ncol=p1)
#print(X2)
print(dim(X2))


pvalue <- c()
for (i in 1:p1){
  df = data.frame(column1 = Y, column2 = X2[,i])
  Test = t.test(df$column2[which(df$column1==1)],
                df$column2[which(df$column1==0)],
                level = 0.95)
  pvalue <- c(pvalue,Test$p.value)
}
pv2 = pvalue
print(summary(pv2))
hist(pv2,20,main="p-value test colonnes de X2 et Y",col = "blue")
print(length(which(pv2 < 0.05))) # number of significant tests
```

The covariates are linked with Y normally. Exactly, we see it because most p-value < 0.05

## c) other matrice X3, and min, max, histogram of p-value's vector and number of Significant tests and the empirical power of the test

```{r}
vect = c()
for (j in 1:p1){
Z <- (runif(n,0,1)<0.7)*1
obs <- Z*rnorm(n,1*Y,1) + (1-Z)*rnorm(n,0,1) #obs : n realisations de X_j
vect = c(vect,obs)
}
X3 <- matrix(vect,ncol=p1)
#print(X3)
print(dim(X3))


pvalue <- c()
for (i in 1:p1){
  df = data.frame(column1 = Y, column2 = X3[,i])
  Test = t.test(df$column2[which(df$column1==1)],
                df$column2[which(df$column1==0)],
                level = 0.95)
  pvalue <- c(pvalue,Test$p.value)
}
pv3 = pvalue
print(summary(pv3))
hist(pv3,20,main="p-value test colonnes de X3 et Y",col = "blue")
print(length(which(pv3 < 0.05))) # number of significant tests
```

The covariates are linked with Y normally. Exactly, we see it because most p-value < 0.05

## d) other matrice X4, and min, max, histogram of p-value's vector and number of Significant tests and the empirical power of the test

```{r}
vect = c()
for (j in 1:p1){
Z <- (runif(n,0,1)<0.3)*1
obs <- Z*rnorm(n,3*Y,1) + (1-Z)*rnorm(n,0,1) #obs : n realisations de X_j
vect = c(vect,obs)
}
X4 <- matrix(vect,ncol=p1)
#print(X4)
print(dim(X4))


pvalue <- c()
for (i in 1:p1){
  df = data.frame(column1 = Y, column2 = X4[,i])
  Test = t.test(df$column2[which(df$column1==1)],
                df$column2[which(df$column1==0)],
                level = 0.95)
  pvalue <- c(pvalue,Test$p.value)
}
pv4 = pvalue
print(summary(pv4))
hist(pv4,20,main="p-value test colonnes de X4 et Y",col = "blue")
print(length(which(pv4 < 0.05))) # number of significant tests
```

The covariates are linked with Y normally. Exactly, we see it because most p-value < 0.05

## e) other matrice X5, and min, max, histogram of p-value's vector and number of Significant tests and the empirical power of the test

```{r}
vect = c()
for (j in 1:p1){
Z <- (runif(n,0,1)<0.3)*1
obs <- Z*rnorm(n,2*Y,1) + (1-Z)*rnorm(n,0,1) #obs : n realisations de X_j
vect = c(vect,obs)
}
X5 <- matrix(vect,ncol=p1)
#print(X5)
print(dim(X5))


pvalue <- c()
for (i in 1:p1){
  df = data.frame(column1 = Y, column2 = X5[,i])
  Test = t.test(df$column2[which(df$column1==1)],
                df$column2[which(df$column1==0)],
                level = 0.95)
  pvalue <- c(pvalue,Test$p.value)
}
pv5 = pvalue
print(summary(pv5))
hist(pv5,20,main="p-value test colonnes de X5 et Y",col = "blue")
print(length(which(pv5 < 0.05))) # number of significant tests
```

The covariates are linked with Y normally. Exactly, we see it because most p-value < 0.05

## f) other matrice X6, and min, max, histogram of p-value's vector and number of Significant tests and the empirical power of the test

```{r}
vect = c()
for (j in 1:p1){
Z <- (runif(n,0,1)<0.3)*1
obs <- Z*rnorm(n,1*Y,1) + (1-Z)*rnorm(n,0,1) #obs : n realisations de X_j
vect = c(vect,obs)
}
X6 <- matrix(vect,ncol=p1)
#print(X6)
print(dim(X6))


pvalue <- c()
for (i in 1:p1){
  df = data.frame(column1 = Y, column2 = X6[,i])
  Test = t.test(df$column2[which(df$column1==1)],
                df$column2[which(df$column1==0)],
                level = 0.95)
  pvalue <- c(pvalue,Test$p.value)
}
pv6 = pvalue
print(summary(pv6))
hist(pv6,20,main="p-value test colonnes de X6 et Y",col = "blue")
print(length(which(pv6 < 0.05))) # number of significant tests
```

## g) Comment the results

le nombre de tests significatifs diminue

## h) for n = 100, and for n = 200

## 9) Compute the vector of the p-values of the 600 tests that test the link between the Xj and Y. Look and the min, the max and do the histplot, comment

```{r}
pv7 = c(pv1,pv2,pv3,pv4,pv5,pv6) 
length(pv7)
print(summary(pv7))
boxplot(pv1,pv2,pv3,pv4,pv5,pv6)
hist(pv7,20,main="p-value test colonnes de X et Y",col = "blue")

```
## 10) Compute the vector of the p-values of the other 600 tests that test the link between the Xj and Y. Look and the min, the max and do the histplot, comment

simulate 600 other covariates independent from Y .

```{r}
p <- 6*p1
X7 <- matrix(rnorm(n*p,0,1), ncol=p)

```



```{r}
pvalue <- c()
for (i in 1:p){
  df = data.frame(column1 = Y, column2 = X7[,i])
  Test = t.test(df$column2[which(df$column1==1)],
                df$column2[which(df$column1==0)],
                level = 0.95)
  pvalue <- c(pvalue,Test$p.value)
}
pv8 = pvalue
summary(pv8)
hist(pv8,20,main="p-value test others colonnes de X et Y",col = "red")
```

## 11) Aggregate the two vectors of the p-values and look and the min, the max and do the histplot,comment

```{r}
pv9 = c(pv7,pv8)
length(pv9)
print(summary(pv9))
boxplot(pv7,pv8, col=c("blue","red"))
hist(pv9,20,main="p-value test all colonnes de X et Y")
#pv9 : all p-values's vector (question 11)
#pv8 : when X_j are not linked with Y (question 10)
#pv7 : whan X_j  are linked with Y (question 9)
```


## 12) Choose a risk alpha for each test and compute each entry of the following table (frequencies associated with the test's decisions). Remember what is fixed and what is dependent of the simulation.

```{r}
S = length(which(pv7 < 0.05))   # vrai positif
print("S : nombre de vrais positifs")
S

V = length(which(pv8 < 0.05))   # faux positifs
print("V : nombre de faux positifs")
V

#R = length(which(pv9 < 0.05))   # positif
R = S+V
print("R : nombre de tests positifs")
R

U = length(which(pv8 > 0.05 ))  # faux négatifs
print("U : faux negatifs")
U

T = length(which(pv7 > 0.05)) # negatif
print("T : nombre de tests negatifs")
T

```

# 3. Adjusted p-values

## 1. a) For each of the method, Bonferroni, Sidak, Holms and Bejamini-Hochberg, compute the vector of adjusted p-values, look at the min, the max, plot the histogram and comment

```{r}
print("bonferroni's method")
adj.pv = p.adjust(pv9, method = "bonferroni")
summary(adj.pv)
hist(adj.pv,20, main = "with benferroni's method ")
Bonferroni = adj.pv

print("Sidak's method")
adj.pv = 1-(1-pv9)**length(pv9)
summary(adj.pv)
hist(adj.pv,20,  main = "with Sidak's method ")
Sidak = adj.pv

print("holm's method")
adj.pv = p.adjust(pv9, method = "holm")
summary(adj.pv)
hist(adj.pv,20, main = " with holm's method ")
Holm = adj.pv

print("BH's method")
adj.pv = p.adjust(pv9, method = "BH")
summary(adj.pv)
hist(adj.pv,20, main = " with BH's method ")
BH = adj.pv

```

## b) Compare the previous methods by plotting the point clouds of each method 2 to 2.

```{r}
pairs(cbind(Bonferroni,Sidak,Holm,BH))
```


## c) after installation package qvalue

```{r}
library(qvalue)

```

## d) Compute the qvalues by the following code and do the table and the comparison

```{r}
pval2 = pv9
qvalcom<-qvalue(pval2)
qval<-qvalcom$qvalues
summary(qval)
hist(qval, 20, main = "q-value")
```
# 2.
```{r}
p = length(pv9)
lambda = 0.25
pi0 = (V+U)/p
print("pi0")
pi0
pi0.estime = length(which(pv9 > lambda))/(p*(1-lambda))
print("pi0 estimee")
pi0.estime
```

# 4. Application to real data

On importe le jeu de données datacancer, il a au départ 143 lignes et 104 colonnes. On fait une analyse pour les types de données.
On constate que beaucoup de colonnes sont de type "character", on les transforme alors
en "factor". On décide d'identifier et de supprimer les colonnes de type factor qui ont qu'une seule modalité. datacancer a maintenant 143 lignes et 98 colonnes.

```{r}
datacancer <- read.csv("~/Documents/MCAD Marcel/Stat de grande dimension/TP1/datacancer.csv")

dim(datacancer)


#str(datacancer)

ind = NULL
for (j in 1: ncol(datacancer)){
  if (is.character(datacancer[,j])){
    datacancer[,j] = factor(datacancer[,j])
  }
  if (length(levels(datacancer[,j]))==1){
    ind = c(ind,j)
  }
}
#print(ind)
datacancer = datacancer[,-ind]
dim(datacancer)


```

On identifie la variable Y comme cible, on a envie de voir les tables de contingence de Y avec chaque variable catégorielle d'au moins 3 modalités pour vérifier les conditions de faisabilité du test de khi-deux qui impose que les effectifs soient supérieurs à 5.

```{r}
Y = factor(datacancer$Surv12)
for (j in 1: ncol(datacancer)){
  if (is.factor(datacancer[,j])){
    if (length(levels(datacancer[,j]))>2){
      print(j)
      print(table(Y,datacancer[,j]))
    }
  }
}
```

On regroupe les modalités de certaines variables catégorielles afin qu'on puisse
avoir les conditions du test de Khi-2 ou du test exact de Fisher vérifiées.

```{r}

levels(datacancer[,12])[c(2,3)] = c("positive_unknow")
levels(datacancer[,13])[c(1,2,3,4)] = c("autres_stages")
levels(datacancer[,19])[c(2,3,4)] = c("autres")
levels(datacancer[,32])[c(1,2,3)] = c("autres")
levels(datacancer[,34])[c(1,2)] = c("autres")
levels(datacancer[,33])[c(1,2,3)] = c("autres")
levels(datacancer[,35])[c(2,3)] = c("autres")
levels(datacancer[,39])[c(1,2)] = c("autres")
levels(datacancer[,42])[c(1,2)] = c("autres")
levels(datacancer[,45])[c(1,2)] = c("autres")
levels(datacancer[,46])[c(1,2)] = c("autres")
levels(datacancer[,49])[c(1,2)] = c("autres")
levels(datacancer[,50])[c(1,2,3)] = c("autres")
levels(datacancer[,51])[c(1,2)] = c("autres")

Y = factor(datacancer$Surv12)
for (j in 1: ncol(datacancer)){
  if (is.factor(datacancer[,j])){
    if (length(levels(datacancer[,j]))>=2){
      print(j)
      print(table(Y,datacancer[,j]))
    }
  }
}

```

On supprime les colonnes 1, 2 qui sont juste des identifiants, on enlève aussi
la colonne 6 et la colonne 7, qui correspondent respectivement à Surv6 et à Surv12, car ce sont  des variables d'intérêts. On enlève aussi les colonnes 5 (censoros) et 8 (censorTRT), de même que les colonnes 3 (TRT) et 4 (OS). On crée un nouveau datacancer
où la première colonne est la variable cible suivie dans l'ordre des covariables qualitatives (au nombre de 46) puis des covariables quantitatives (au nombre de 44).

```{r}
datacancer = datacancer[-c(1,2,3,4,5,6,7,8)]

ind_qual <- sapply(datacancer, function(x) is.factor(x))
qual <- datacancer[ ,ind_qual]

ind_quant <- sapply(datacancer, function(x) is.numeric(x))
quant <- datacancer[ ,ind_quant]

datacancer = cbind(Y,qual,quant)
dim(datacancer)
```

### 1) Select the qualitative covariates and perform the appropriate tests to see the link with Surv12. Save the p-values in a vector.

On fait un test exact de Fisher dans le cas où la table de contingence est de taille 2x2 et au moins un effectif est plus petit que 5.
On fait un test de Khi-deux dans le cas où la table de contingence a tous les effectfs supérieurs à 5.
Lorsque la covariable a au moins 3 modalités, on regroupe certaines modalités si on ne peut pas faire (sans erreur) le test de Khi-deux

```{r}

pval.qual = NULL
for(j in 1:ncol(qual)){
  tab = table(Y,qual[,j])
  if (length(levels(qual[,j]))==2){
    if ((tab[1,1]<=5 || tab[1,2]<=5) ||( tab[2,1]<=5 || tab[2,2]<=5)){
      test = fisher.test(Y,qual[,j])
      pval.qual = c(pval.qual,test$p.value)
    }
    else{
      test = summary(tab)
      pval.qual = c(pval.qual,test$p.value)
    }
  }
  else{
    test = summary(tab)
      pval.qual = c(pval.qual,test$p.value)
  }
}
print(summary(pval.qual))
hist(pval.qual,20,main="p-value test de lien entre Y et les X_j catégorielles",
     col = "yellow")
```

### 2 ) Select the quantitative covariates and perform the appropriate tests to see the link with Surv12. Save the p-values in a vector

On fait un test de comparaison de deux moyennes. 

```{r}
pval.quant = NULL
Y = as.numeric(Y)
for (j in 1:ncol(quant)){
  test = t.test(Y,quant[,j])
  pval.quant = c(pval.quant,test$p.value)
}
print(summary(pval.quant))
hist(pval.quant,20,main="p-value test de lien entre Y et les X_j quantitatives",
     col = "green")
```

### 3) Aggregate the two vectors and adjust the p-values with different methods. Gives the variables that can explain the survival.

On ajuste les p-valeurs (on les augmente) car on fait des tests multiples. Quelque soit la méthode ("holm", "hochberg", "hommel", "bonferroni", "BH", "BY", etc), les tests significatifs sont ceux dont la p-value ajustée est inférieur à 0.05.
On crée un code pour récupérer les variables significatives (p-value ajustée < 0.05)

```{r}
print("BH's method")
adj.pv = p.adjust(c(pval.qual,pval.quant), method = "BH")
hist(adj.pv,20, main = "with BH's method ")
summary(adj.pv)
indice_variable_expl = which(adj.pv < 0.05)+1
datacancer2 = datacancer[c(1,indice_variable_expl)]
dim(datacancer2)
colnames(datacancer2)
# sauvegarde du dataset avec Surv12 et les variables selectionnees
write.csv2(datacancer2, file ="datacancer2.csv")
```
